---
title: "Document supporting `PCR duplicate proportion estimation and consequences for DNA copy number calculations'"
author: "Andy Lynch"
date: "16/01/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Duplicates
This is a minimal document to generate the results in the manuscript. 

```{r prelims}
BiocManager::install("BSgenome.Hsapiens.UCSC.hg19")
library("Rsamtools")
library("FragmentationDuplicates")
EDpath<-system.file("extdata", package="FragmentationDuplicates")
```

To inform our work, we applied the tool `Picard' to various different regions of the genome and summarize the results in this table. It isn't presented in the paper, but is the basis for some discussion at the end of part 1.1.

```{r picard}
unmaskedP<-read.delim(paste(EDpath,"/Picard/samples.metrics.txt",sep=""),as.is=T)

readPairsExamined <- readPairDuplicates <- OpticalPairDuplicates <- matrix(ncol=9, nrow=22, dimnames = list(unmaskedP[,1], c("Total","X","Y","M","Centromeres","Telomeres","LowCov","HighCov","Residual")))

readPairsExamined[,1]<-unmaskedP[,3]
readPairDuplicates[,1]<-unmaskedP[,6]
OpticalPairDuplicates[,1]<-unmaskedP[,7]

WeaverSuppTable1<-read.delim(paste(EDpath,"/WeaverSuppTable1.txt",sep=""),sep="",header=T,as.is=T)
WeaverSuppTable1$CaseID<-WeaverSuppTable1$CaseID-1
#WeaverSuppTable9<-WeaverSuppTable9[match(WeaverSuppTable1$CaseID, WeaverSuppTable9$CaseID),]

Group<-1+2*(WeaverSuppTable1[,2]=="Male")+(WeaverSuppTable1[,12]=="Blood")
GroupN<-c("Female_Tissue","Female_Blood","Male_Tissue","Male_Blood")[Group]
GroupT<-c("Tissue","Blood","Tissue","Blood")[Group]

maskFiles <- list.files(paste(EDpath,"/masks/",sep=""), full.names = TRUE)
MASKS <- lapply(maskFiles, read.delim, header=FALSE, as.is=TRUE, skip=1)
EGS<-rep(0,9)
EGS[1]<-sum(as.numeric(MASKS[[5]][17:38,3]))+MASKS[[1]][1,3]+MASKS[[2]][1,3]+MASKS[[3]][1,3]
for(i in 1:7) {EGS[i+1] <- sum(MASKS[[i]][,3] - MASKS[[i]][,2])}
EGS[9]<-EGS[1]-sum(EGS[2:8])

metfiles<-list.files(paste(EDpath,"/Picard/",sep=""))
metfiles<-grep("mask",metfiles,value=T)

for(i in 1:7){
  temp<-read.delim(paste(EDpath,"/Picard/",metfiles[i],sep=""),as.is=T)
  temp<-temp[-(15:19),]
  readPairsExamined[,i+1]<-temp[,3]
  readPairDuplicates[,i+1]<-temp[,6]
  OpticalPairDuplicates[,i+1]<-temp[,7]
}

readPairDuplicates[,9]<-readPairDuplicates[,1]-apply(readPairDuplicates[,2:8],1,sum)
readPairsExamined[,9]<-readPairsExamined[,1]-apply(readPairsExamined[,2:8],1,sum)
OpticalPairDuplicates[,9]<-OpticalPairDuplicates[,1]-apply(OpticalPairDuplicates[,2:8],1,sum)

reptable<-matrix(0,ncol=4,nrow=8)
reptable[1,]<-round(100*sapply(split(((readPairDuplicates[,9]/readPairsExamined[,9])),Group),mean),2)
for(i in 2:8){
  reptable[i,]<-round(sapply(split(((readPairDuplicates[,i]/readPairsExamined[,i])/(readPairDuplicates[,9]/readPairsExamined[,9])),Group),mean),2)
}
#reptable

ResDupR<-readPairDuplicates[,9]/readPairsExamined[,9]
ResDepth<-round(200*(readPairsExamined[,9]-OpticalPairDuplicates[,9])/EGS[9],2)

ResDepthDupRem<-round(200*(readPairsExamined[,9]-readPairDuplicates[,9]-OpticalPairDuplicates[,9])/EGS[9],2)

  BasicEst<-(readPairDuplicates[,1]-OpticalPairDuplicates[,1])/
  (readPairsExamined[,1]-OpticalPairDuplicates[,1])
  ResidEst<-(readPairDuplicates[,9]-OpticalPairDuplicates[,9])/
  (readPairsExamined[,9]-OpticalPairDuplicates[,9])


```


Summaries of the heterozygous SNPs from the example data are given in the package, and we process them here using the methods presented in the paper.
```{r genestimates}

HetSNPTable<-read.delim(file.path(EDpath,"HetSNPDups.txt"),as.is=T)
PL<-genPL(16)
CL<-genCoefs(PL)
FDres<-processduptable(HetSNPTable,PL,CL)
```

We now produce figure 2 of the paper.

```{r estimateplot}

pdf("estimates.pdf",width=12.2,height=6)

  par(mfrow=c(1,2))
par(mar=c(5.1,5.1,4.1,1.1))
plot(c(0,0.15),c(0,0.15),type="n",xlab="Proportion from SNPs",ylab="",main="Uncorrected Proportions",las=1,cex.lab=1.5,cex.axis=1.3,cex.main=1.5)
for(i in 1:22){
lines(c(FDres[i,1]/100,FDres[i,1]/100),c(BasicEst[i],ResidEst[i]))
points(FDres[i,1]/100,BasicEst[i],pch=1)
points(FDres[i,1]/100,ResidEst[i],pch=17,col="black")
}
mtext("Genome-wide Proportions",2,3.5,cex=1.5)
abline(0,1)
legend("bottomright",pch=c(1,17),legend=c("Basic Proportion","Masked Proportion"),cex=1.5)
#legend("bottomright",fill=c("red","black"),legend=c("Basic Estimate","Masked Estimate"))

plot(100*BasicEst/FDres[,3],FDres[,1]/FDres[,3],xlim=c(0.9,2.6),ylim=c(0.9,1.3),pch=16,xlab="Basic Proportion/Corrected Estimate",ylab="Observation from SNPs/Corrected Estimate",main="PCR duplicate rate",las=1,cex.lab=1.5,cex.axis=1.3,cex.main=1.5)
abline(v=1,lwd=2)
abline(h=1,lwd=2)
dev.off()
```

Now we perform the processing necessary to produce figure 3 of the paper.

```{r singleend}
HetSNPTableSE<-read.delim(file.path(EDpath,"HetSNPTableSingleEnd.tsv"),as.is=T,header=T)
SEests<-processduptable(HetSNPTableSE,PL,CL,reso=1001)
head(SEests)

pdf("singleend.pdf",width=12.2,height=6)

par(mfrow=c(1,2))
plot(FDres[,1],SEests[,1],pch=16,xlab="Basic estimate from PE data (%age)",ylab="Basic estimate from SE data (%age)",xlim=c(0,40),ylim=c(0,40),las=1,cex.lab=1.5,cex.axis=1.3,cex.main=1.5)
abline(0,1)
plot(FDres[,3],SEests[,3],pch=16,xlab="Corrected estimate from PE data (%age)",ylab="Corrected estimate from SE data (%age)",xlim=c(0,12),ylim=c(0,12),las=1,cex.lab=1.5,cex.axis=1.3,cex.main=1.5)
abline(0,1)
dev.off()
```

This section draws on summaries derived for different copy number states and then produces Figure 4.
```{r tumour}
HetSNPTableAABB<-read.delim(file.path(EDpath,"HetSNPTableTumourAABB.tsv"),as.is=T,header=T)
HetSNPTableAB<-read.delim(file.path(EDpath,"HetSNPTableTumourAB.tsv"),as.is=T,header=T)
HetSNPTableTumour<-rbind(HetSNPTableAABB,HetSNPTableAB)
rownames(HetSNPTableTumour)<-c("AABB","AB")
rm(HetSNPTableAB,HetSNPTableAABB)
tumourresults<-processduptable(HetSNPTableTumour,PL,CL)
```

Here we generate data for Figure 7 in the paper:

```{r mito}
i<-3
lf<-read.delim(paste(EDpath,"/Picard/",metfiles[i],sep=""),as.is=T)
lf<-lf[-(15:19),]

#With all duplicates retained

NuclearCovPerCopyRet<-ResDepth/2
MitoCovDupRet<-200*lf[,3]/16569

MTCNdupret<-MitoCovDupRet/NuclearCovPerCopyRet

#With duplicates removed

NuclearCovPerCopyRem<-ResDepthDupRem/2
MitoCovDupRem<-200*(lf[,3]-lf[,6]-lf[,7])/16569

MTCNduprem<-MitoCovDupRem/NuclearCovPerCopyRem

#Corrected


NuclearCovPerCopy<-(100-FDres[,3])*ResDepthDupRem/(200-2*FDres[,1])
MitoCov<-200*lf[,3]*(1-FDres[,3]/100)/16569

MTCNcor<-MitoCov/NuclearCovPerCopy

  
```

These are the data from Yaun et al. which is citation 15 in the paper.

```{r PCAWG}

## From supplementary material for Yuan et al (2020). Comprehensive molecular characterization of mitochondrial genomes in human cancers. Nature Genetics, 52(3), 342–352.
PCAWG<-c(300.6549982,130.610943,603.9354818,318.8582608,247.1975181,75.11459474,195.4605537,101.4383611,267.9739402,177.7428225,320.8641443,454.414171,193.4416843,183.1840712,243.3734567,371.5073704,137.2701945,196.9313568,225.4146729,138.5534034,279.5720601,164.4498631,140.5688875,156.9191376,143.5136508,310.5046524,187.2813062,146.9098841,147.2362958,141.8160834,111.4595505,145.5640682,120.2311169,150.7269393,166.5555356,126.5012388,211.8039265,198.5621459,134.3952882,144.6492944,102.1598613,172.2856122,230.4994119,216.8637284,162.4593709,320.5602413,134.6479081,75.39691491,147.0414197,402.4110327,116.2298472,156.3966169,302.9745917,532.5634505,150.6302249,201.076587,118.7878462,135.6954622,74.27481745,253.4217658,249.13731,237.0978902,120.4590114,110.9118497)
```

and now we produce figure 7.

```{r mitoplot}
pdf("mitoCN.pdf",width=12.2,height=5)
par(mar=c(5.1,4.6,3.1,.6))
boxplot(MTCNcor[GroupT=="Tissue"],MTCNcor[GroupT=="Blood"],MTCNduprem[GroupT=="Tissue"],MTCNduprem[GroupT=="Blood"],ylim=c(0,800),PCAWG,xaxs="i",yaxs="i",las=2,axes=F,cex.lab=1.5,cex.axis=1.3)
axis(2,las=2)
box()
abline(v=2.5,lty=2,lwd=3)
abline(v=4.5,lwd=3)
mtext("Corrected",3,at = 1.5,cex=1.5,line=1)
mtext("All Duplicates Removed",3,at = 3.5,cex=1.5,line=1)
mtext("PCAWG",3,at = 5,cex=1.5,line=1)
mtext("Estimated mtDNA copy number",2,at=400,cex=1.5,line=3)
axis(1,at=1:5,labels=c("Benign\nTissue","Blood","Benign\nTissue","Blood","Tumour\ntissue"),tick=F,cex.axis=1.5,line=2)
dev.off()
```

and another version

```{r mitoplot2}
pdf("mitoCNalt.pdf",width=12.2,height=5)
par(mar=c(5.1,4.6,3.1,.6))
boxplot(MTCNdupret[GroupT=="Tissue"],MTCNdupret[GroupT=="Blood"],MTCNduprem[GroupT=="Tissue"],MTCNduprem[GroupT=="Blood"],MTCNcor[GroupT=="Tissue"],MTCNcor[GroupT=="Blood"],ylim=c(0,800),PCAWG,xaxs="i",yaxs="i",las=2,axes=F,cex.lab=1.5,cex.axis=1.3)
axis(2,las=2)
box()
abline(v=2.5,lty=2,lwd=3)
abline(v=4.5,lty=2,lwd=3)
abline(v=6.5,lwd=3)
mtext("Duplicates Retained",3,at = 1.5,cex=1.5,line=1)
mtext("Corrected",3,at = 5.5,cex=1.5,line=1)
mtext("All Duplicates Removed",3,at = 3.5,cex=1.5,line=1)
mtext("PCAWG",3,at = 7,cex=1.5,line=1)
mtext("Estimated mtDNA copy number",2,at=400,cex=1.5,line=3)
axis(1,at=1:7,labels=c("Benign\nTissue","Blood","Benign\nTissue","Blood","Benign\nTissue","Blood","Tumour\ntissue"),tick=F,cex.axis=1.5,line=2)
dev.off()
```

Here we define some functions for generating some simulated reads. Note that the allele-specific aspects (defining A and B copy numbers separately) are not used.
```{r Simulations}


simreads<-function(A=1,B=1,cov=Coverage,GL=GenomeLength,
                   ISM=InsertSizeM,ISS=InsertSizeS,RL=ReadLength){
  Areads<-GL*cov*A/(2*RL)
  Breads<-GL*cov*B/(2*RL)  
  
  Astarts<-sample(GL,Areads,replace=T)
  Aends<-Astarts+2*RL+round(pmax(rnorm(Areads,ISM,ISS),0))
  
  Bstarts<-sample(GL,Breads,replace=T)
  Bends<-Bstarts+2*RL+round(pmax(rnorm(Breads,ISM,ISS),0))
  
  reorder<-sample(Areads+Breads)
  
  starts=c(Astarts,Bstarts)[reorder]
  ends=c(Aends,Bends)[reorder]
  Group=rep(c("A","B"),c(Areads,Breads))[reorder]
  return(list(starts=starts,ends=ends,group=Group))
}

markdups<-function(mylist){
  key<-paste(mylist$starts,mylist$ends,sep="-")
  mylist$duplicates=duplicated(key)
  return(mylist)
}



getobs<-function(CN,cov=Coverage,GL=GenomeLength,
                 ISM=InsertSizeM,ISS=InsertSizeS,RL=ReadLength){
  tmp<-simreads(A=CN/2,B=CN/2,cov=cov,GL=GL,ISM=ISM,ISS=ISS,RL=RL)
  tmp<-markdups(tmp)
  return(sum(!tmp$duplicates)*2*ReadLength/GenomeLength)
}

```

The work for figure 6 is now presented.

```{r}
xvals<-c(2,seq(50,1000,50))

GenomeLength<-100000
InsertSizeM<-300
InsertSizeS<-40
ReadLength<-100
Coverage=30
SNPspace<-500



yvals<-rep(NA,21)
for(i in 1:21){
  yvals[i]<-getobs(xvals[i])
}

GenomeLength<-100000
InsertSizeM<-300
InsertSizeS<-100
ReadLength<-150
Coverage=15
SNPspace<-500


yvals2<-rep(NA,21)
for(i in 1:21){
  yvals2[i]<-getobs(xvals[i],cov=Coverage,GL=GenomeLength,
                    ISM=InsertSizeM,ISS=InsertSizeS,RL=ReadLength)
}



GenomeLength<-100000
InsertSizeM<-300
InsertSizeS<-40
ReadLength<-75
Coverage=30
SNPspace<-500


yvals3<-rep(NA,21)
for(i in 1:21){
  yvals3[i]<-getobs(xvals[i],cov=Coverage,GL=GenomeLength,
                    ISM=InsertSizeM,ISS=InsertSizeS,RL=ReadLength)
}


pdf("simmtDNACNB.pdf",width=12.2,height=5)
par(mar=c(5.1,4.6,1.1,.6))
plot(xvals,100-(200*yvals/yvals[1])/xvals,type="l",xlim=c(0,1000),ylim=c(0,50),ylab="Percentage Discarded",las=1,xlab="True Copy Number",lwd=3,cex.lab=1.5,cex.axis=1.3,yaxs="i")
lines(xvals,100-(200*yvals2/yvals2[1])/xvals,lwd=3,lty="45")
lines(xvals,100-(200*yvals3/yvals3[1])/xvals,lty="221226",lwd=3)
legend("topleft",legend=c("RL=75, SD=40, Cov=30","RL=100, SD=50, Cov=30","RL=150, SD=100, Cov=15"),lty=c("221226","solid","45"),lwd=3,cex=1.5)
dev.off()
```

Finally, we look at the work for figure 5. 

```{r}
tumMTCN<-1000
benMTCN<-500
tumCN<-3
NucGenome<-3000000000
MTGenome<-16569

Tumprop<-0.9
InsertSizeM<-300
InsertSizeS<-40
ReadLength<-100


result<-matrix(NA,ncol=14,nrow=11)

for(i in 1:14){

tumMTCN<-seq(200,1500,100)[i]

for(j in 1:11){

Tumprop<-seq(0.5,1,0.05)[j]


Noreads<-1000000000

Nobases<-Noreads*2*ReadLength

TotGen<-NucGenome*(tumCN*Tumprop+2*(1-Tumprop))+MTGenome*(tumMTCN*Tumprop+benMTCN*(1-Tumprop))

Cov<-Nobases/TotGen

mtsim<-simreads(A=1,B=(tumMTCN*Tumprop+(1-Tumprop)*benMTCN)-1,cov=Cov,GL=MTGenome,ISM=InsertSizeM,ISS=InsertSizeS,RL=ReadLength)
mtsim<-markdups(mtsim)


nucsim<-simreads(A=1,B=tumCN*Tumprop+1-2*Tumprop,cov=Cov,GL=1000000,ISM=InsertSizeM,ISS=InsertSizeS,RL=ReadLength)
nucsim<-markdups(nucsim)

result[j,i]<-table(mtsim$duplicates)["FALSE"]*2*ReadLength*prop.table(table(nucsim$duplicates))["FALSE"]/(MTGenome*Cov)
}}


pdf("mtDNAbiases.pdf",width=12.2,height=5)
par(mar=c(5.1,4.6,3.1,.6))
plot(seq(200,1500,100),result[1,],type="l",ylim=c(200,1500),xlim=c(200,1600),xlab="True tumour mtDNA copy number",ylab="Inferred copy number")
abline(0,1,lwd=3)
lines(seq(200,1500,100),result[11,],type="l",ylim=c(200,1500))
lines(seq(200,1500,100),result[6,],type="l",ylim=c(200,1500))
abline(v=500,lty=2)
text(1550,660,"50%")
text(1550,786,"75%")
text(1550,908,"100%")
dev.off()
```

